---
title: "External CITEseq dataset analysis"
author: "Gillian Savage"
date: "1/7/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
#make sure you are using seurat V4
#can install Seurat v4 as follows:
#remotes::install_version("SeuratObject", "4.1.4", repos = c("https://satijalab.r-universe.dev", getOption("repos")))
#remotes::install_version("Seurat", "4.4.0", repos = c("https://satijalab.r-universe.dev", getOption("repos")))
library(tidyverse)
library(RColorBrewer)
library(ggrepel)
library(ggpubr)
library(rstatix)
library(patchwork)
library(R.utils)
library(igraph)
library(dittoSeq)
library(ggsci)
library(hues)
library(pheatmap)

mutate <- dplyr::mutate
filter <- dplyr::filter
count <- dplyr::count
summarise <- dplyr::summarise
rename <- dplyr::rename
select <- dplyr::select
options(dplyr.summarise.inform=FALSE)

```


Counts data was obtained from the European Genome Archive ([EGAD50000000497](https://ega-archive.org/datasets/EGAD50000000497)). Data was processed and integrated using workflow described in Roider T, Baertsch MA, Fitzgerald D, Vöhringer H, Brinkmann BJ, Czernilofsky F, et al. Multimodal and spatially resolved profiling identifies distinct patterns of T cell infiltration in nodal B cell lymphoma entities. Nat Cell Biol. 2024;26:478–89. 


# Read in processed data
```{r}
Combined_T <- readRDS("~/trms_dlbcl/external_data/Tcells_Integrated.rds") 
```

# UMAP visualizations

UMAP of clusters
```{r fig.height=5, fig.width=5.5}
DimPlot(Combined_T, reduction = 'wnn.umap', label = T, raster = T)+
  NoLegend()
```

```{r}
#remove singeltons
Combined_T <- subset(Combined_T, idents = c(0:16))
Combined_T$IdentI <- Idents(Combined_T)
```

Replot UMAP and color for Figure 5A
```{r fig.height=5, fig.width=5.5}
DimPlot(Combined_T, reduction = 'wnn.umap', label = T, cols = c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#6A6767","#EE4C97","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"), raster = T)+
  NoLegend()
```
Plot Lymphoma entity for Figure 5B
```{r}
Combined_T <- SetIdent(Combined_T, value = "Entity")

DimPlot(Combined_T, reduction = 'wnn.umap', label = F, cols = c("#7876B1", "#E18727", "#20854E", "#0072B5", "#BC3C29"), raster = T)

```

Plot CD103 expression on UMAP as detected by antibody capture (protein expression of CD103) for Figure 5C. Cells positive for CD103 determined by cutoff value of 1.5.

```{r fig.height=5, fig.width=5.5}
#convert to ADT assay
DefaultAssay(Combined_T) <- "integratedADT"
# create new column in the metadata for Neg/Pos status of CD103
Combined_T@meta.data$CD103 <- 'Neg'
Combined_T@meta.data$CD103[which(Combined_T@assays$integratedADT@data[".CD103",]>1.5)] <- 'Pos'

#Set indent to CD103
Combined_T <- SetIdent(Combined_T, value = "CD103")

#generate UMAP showing positive cells (Fig. 5C)
DimPlot(Combined_T, reduction = 'wnn.umap', label = F, cols = c("Neg" = "grey", "Pos" = "#DC0000B2"))
```

Graph the proportion of CD103+ T cells per lymphoma entity (Fig. 5H)
```{r}
my_comparisons <- list(c("FL", "MCL")) #only one found to be significant
dittoFreqPlot(Combined_T, "CD103",
    sample.by = "subject_id", group.by = "Entity", vars.use = "Pos",color.panel = c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1"))+
# Add pairwise comparisons p-value
stat_compare_means(comparisons = my_comparisons, label.y = c(0.08, 0.11))
```

# Identifying a Trm cluster

Look at Trm score within each cluster:
```{r}
#switch to RNA assay
DefaultAssay(Combined_T) <- "RNA"
#load genes associated with Trm sig
Trm.genes <- c("ITGAE",	"AHR",	"ANXA1",	"CTLA4",	"CXCR6",	"DUSP5",	"EPAS1",	"GZMB",	"IFNG",	"ITGAV",	"KLRC1",	"LMNA",	"RBPJ",	"TNFRSF9",	"XCL1",	"PDCD1",	"DUSP6",	"IL10",	"IL2",	"KCNK5",	"CXCL13",	"ITGA1",	"RGS1", "PRDM1", "RUNX3", "CD69", "DAPK2")

#add score for each cell for tissue residency
Combined_T <- AddModuleScore(Combined_T,
                  features = list(Trm.genes),
                  name="Trm.score")

#pull out metatdata
Combined_T.df <- data.frame(Combined_T@meta.data)

```

Plot Trm score per cell on UMAP (Fig. 5D)
```{r fig.height=5, fig.width=5.5}
#plot on UMAP
FeaturePlot(Combined_T,reduction = 'wnn.umap',
            features = "Trm.score1", label = TRUE, repel = TRUE) +
            scale_colour_gradientn(colours = brewer.pal(n = 11, name = "YlOrRd"))
```

Trm Score per cluster (not shown in manuscript)
```{r}
ggplot(Combined_T.df,aes(x=IdentI,y=Trm.score1,fill=IdentI))+geom_violin()+theme_classic()+ggtitle("Tissue Residency score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+
  scale_fill_manual(values=c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#6A6767","#EE4C97","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"))
```

Cluster 8 has highest TRM score and also expresses CD103 at the protein level.

To further confirm if cluster 8 is a Trm cell we look at known TRM genes by heatmap.
```{r}
# make a empty data table
corrected.cluster.means <- data.table()


# seurat_clusters is the column where you store the cluster information
# this part will calculate the average expression of genes in each cell cluster
for (i.clust in levels(Combined_T@meta.data$IdentI)) {
	i.clust.means <- data.frame(rowMeans(Combined_T[["RNA"]]@data[,which(Combined_T@meta.data$IdentI== i.clust)]))
	colnames(i.clust.means) <- i.clust
	rownames(i.clust.means) <- rownames(Combined_T[["RNA"]]@data)
	if (nrow(corrected.cluster.means) < 1) {
		corrected.cluster.means <- i.clust.means
	}
	else {
		corrected.cluster.means <- cbind(corrected.cluster.means, i.clust.means)
	}
}
```


```{r}
# read the markers list
comp.func.data <- read.delim("~/trms_dlbcl/internal_data/TRM_heatmap.txt", stringsAsFactors = FALSE)

comp.func.data$Component <- as.factor(comp.func.data$Component)

# gene annotation 
comp.func.ann <- comp.func.data[,"Component", drop = FALSE]
rownames(comp.func.ann) <- comp.func.data$Gene
# assign colors for the genes
comp.func.ann.colors <- list(Component = iwanthue(nlevels(comp.func.data$Component)))
names(comp.func.ann.colors$Component) <- levels(comp.func.data$Component)
comp.func.rownames <- apply(comp.func.data, 1, function(x) {if (x["Alt_name"] == "") { return(x["Gene"]) } else { return(paste0(x["Gene"], " (", x["Alt_name"], ")")) }})

# gaps in the heatmap
gaps.row <- which(!duplicated(comp.func.ann$Component))[-1] - 1

# subset the matrix to keep only the markers
corrected.comp.mat <- as.matrix(corrected.cluster.means[comp.func.data$Gene,])


```

Plot heatmap for Figure 5E
```{r}
(pheatmap(corrected.comp.mat, 
          scale = "row", 
          cluster_rows = FALSE, 
          cluster_cols = TRUE, 
          labels_row = comp.func.rownames,
					annotation_colors = comp.func.ann.colors, 
					annotation_row = comp.func.ann, 
					annotation_names_row = FALSE,
					gaps_row = gaps.row,
          color = colorRampPalette(c("navy", "white", "red"))(50)))

```

Graph the proportion of each cluster per lymphoma entity (Figure 5I)
```{r}
dittoBarPlot(Combined_T, "IdentI",group.by = "Entity", color.panel = c("#BC3C29","#0072B5", "#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#6A6767","#EE4C97","#da6857")) 

```

## Analysis of Cluster 8 vs other CD103+ clusters

Cluster 8 had the highest TRM signature score but there were other clusters that expressed CD103 at the protein level - Cluster 7 and Cluster 13. To confirm that cluster 8 is the Trm cluster and explore if cluster 7 and 13 could also be Trm clusters we compared expression of genes and proteins that makes these clusters distinct.

First we compared the proliferation, activation, cytotoxicity, and exhaustion scores of these CD103+ T cell clusters.

```{r}
#load genes to be used for proliferation scoring

proliferation.genes <- c("ZWINT", "E2F1", "FEN1", "FOXM1", "H2AFZ", "HMGB2", "MCM2", "MCM3", "MCM4", "MCM5", "MCM6", "MKI67", "MYBL2", "PCNA", "CCND1", "AURKA", "BUB1", "TOP2A", "TYMS", "DEK", "CCNB1")

#load genes to be used for cytoxicity scoring

cytotox.genes <- c("KLRF1", "GNLY", "CTSW", "NKG7", "KLRD1", "GZMA", "ADGRG1", "CST7", "KLRK1", "FASLG", "HCST", "KLRB1", "ITGB1", "GZMB", "PRF1")

#load genes to be used for exhaustion scoring

exhaustion.genes <- c("HAVCR2", "LAG3", "TIGIT", "CTLA4", "PDCD1", "LAYN", "BTLA", "CD160")

#load genes to be used for activation scoring

activation.genes <- c("IL26", "IL13", "PARP8", "CD63", "IFNG", "GZMB", "STAT1", "TNF", "HLA-DRA", "HLA-DRB5", "HLA-DRB1")

```

```{r}
#Separate out CD103+ clusters
CD103.T <- subset(Combined_T, idents = c("13", "8", "7"))
DefaultAssay(CD103.T) <- "RNA"

#add score for each cell for proliferation
CD103.T <- AddModuleScore(CD103.T,
                  features = list(proliferation.genes),
                  name="Prolif.score")

#add score for each cell for cytoxicity
CD103.T <- AddModuleScore(CD103.T,
                  features = list(cytotox.genes),
                  name="Cytotox.score")

#add score for each cell for exhaustion
CD103.T <- AddModuleScore(CD103.T,
                  features = list(exhaustion.genes),
                  name="EX.score")

#add score for each cell for activation
CD103.T <- AddModuleScore(CD103.T,
                  features = list(activation.genes),
                  name="Activation.score")

#pull out metatdata
CD103.T.df <- data.frame(CD103.T@meta.data)

#subset for scores
CD103.T.df.scores <- CD103.T.df[, c("IdentI", "Prolif.score1", "Cytotox.score1", "EX.score1", "Activation.score1")]

head(CD103.T.df.scores)

```

Plot violin plots of each of these scores between clusters 7, 8 and 13 (Fig. 5F)
```{r}
#proliferation
ggplot(CD103.T.df.scores,aes(x=IdentI,y=Prolif.score1,fill=IdentI))+geom_violin()+theme_classic()+ggtitle("Proliferation score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#6A6767","#EE4C97", "#A8A7CD"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-1, 2.2) + 
  #stat_compare_means(method = "anova", label.y = 1.5)+        # Add global annova p-value
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("7", "8"), c("13", "8")),
                      hide.ns = TRUE, label.y = 1, size = 3)  

#activation

ggplot(CD103.T.df.scores,aes(x=IdentI,y=Activation.score1,fill=IdentI))+geom_violin()+theme_classic()+ggtitle("Activation score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#6A6767","#EE4C97", "#A8A7CD"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-1, 2.2) + 
  stat_compare_means(method = "anova", label.y = 2)+        # Add global annova p-value
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("7", "8"), c("13", "8")),
                      hide.ns = TRUE, label.y = 1, size = 3)  


#cytotoxicity

ggplot(CD103.T.df.scores,aes(x=IdentI,y=Cytotox.score1,fill=IdentI))+geom_violin()+theme_classic()+ggtitle("Cytotoxicity score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#6A6767","#EE4C97", "#A8A7CD"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-1, 2.2) + 
 # stat_compare_means(method = "anova", label.y = 2.2)+        # Add global annova p-value
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("7", "8"), c("13", "8")),
                      hide.ns = TRUE, label.y = 2, size = 3)



#exhaustion

ggplot(CD103.T.df.scores,aes(x=IdentI,y=EX.score1,fill=IdentI))+geom_violin()+theme_classic()+ggtitle("Exhaustion score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#6A6767","#EE4C97", "#A8A7CD"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-1, 2.2) + 
  #stat_compare_means(method = "anova", label.y = 2)+        # Add global annova p-value
   stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("7", "8"), c("13", "8")),
                      hide.ns = TRUE, label.y = 1.6, size = 3)  

```

#Look at TGFb signaling between clusters
```{r}
#load pathways
TCR.sig <- getGmt("~/trms_dlbcl/internal_data/KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY.gmt")
TGF.sig <- getGmt("~/trms_dlbcl/internal_data/KEGG_TGF_BETA_SIGNALING_PATHWAY.gmt")

# Extract gene sets into a named list
tcr_list <- lapply(TCR.sig, geneIds)
names(tcr_list) <- sapply(TCR.sig, function(x) x@setName)

tgf_list <- lapply(TGF.sig, geneIds)
names(tgf_list) <- sapply(TGF.sig, function(x) x@setName)

TCR.sig <- list(tcr_list[["KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY"]])
TGF.sig <- list(tgf_list[["KEGG_TGF_BETA_SIGNALING_PATHWAY"]])
```
Enrichment score calculations:
```{r}
DefaultAssay(Combined_T) <- "RNA"
#add score for each cell for TCR and TGFb signaling
CD103.T <- AddModuleScore(CD103.T,
                  features = TCR.sig,
                  name="TCR.score")

CD103.T <- AddModuleScore(CD103.T,
                  features = TGF.sig,
                  name="TGF.score")


#pull out metatdata
CD103.T.df.scores <- data.frame(CD103.T@meta.data)

#subset for scores
CD103.T.df.scores <- CD103.T.df.scores[, c("IdentI", "CD103", "TCR.score1", "TGF.score1", "Entity")]

head(TCRsig.df)

```

Graphs for Figure 5G
```{r}

ggplot(CD103.T.df.scores,aes(x=IdentI,y=TCR.score1,fill=IdentI))+geom_violin(position = "dodge")+theme_classic()+ggtitle("TCR signaling pathway score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#6A6767","#EE4C97", "#A8A7CD"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.2) + 
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("7", "8"), c("13", "8")),
                      hide.ns = TRUE, label.y = 0.4, size = 3)  

ggplot(CD103.T.df.scores,aes(x=IdentI,y=TGF.score1,fill=IdentI))+geom_violin(position = "dodge")+theme_classic()+ggtitle("TGFb signaling pathway score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#6A6767","#EE4C97", "#A8A7CD"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.2) + 
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("7", "8"), c("13", "8")),
                      hide.ns = TRUE, label.y = 0.2, size = 3)  

```

#Explore gene and protein differences between clusters expressing CD103 at protein level

Look at differences of genes of interest by violin plot (Supp Fig. S8A)
```{r}
VlnPlot(object = CD103.T, features = c('ITGAE', "CXCR6", "PDCD1", "TIGIT", "LAG3", "TOX", "HAVCR2", "GZMB", "GZMA", "PRF1", "IFNG", "CCL3", "CCL4", "CCL5", "SELL", "IL7R", "KLF2", "TCF7", "FOSB"), pt.size = F, cols = c("#6A6767","#EE4C97", "#A8A7CD"))
```

Look at differences of proteins of interest by violin plot (Supp Fig. S8B)
```{r}
DefaultAssay(CD103.T) <- "integratedADT"
VlnPlot(object = CD103.T, features = c('ITGAE', "CXCR6", "PDCD1", "TIGIT", "LAG3", "TOX", "HAVCR2", "GZMB", "GZMA", "PRF1", "IFNG", "CCL3", "CCL4", "CCL5", "SELL", "IL7R", "KLF2", "TCF7", "FOSB"), pt.size = F, cols = c("#6A6767","#EE4C97", "#A8A7CD"))
```

#Specific cluster 7 analysis based on cells in that cluster that are CD103+ vs CD103-
```{r}
#subset cluster 7
Combined_T <- SetIdent(Combined_T, value = "IdentI")
Clus7 <- subset(Combined_T, idents = "7")
#Set indent to CD103
Clus7 <- SetIdent(Clus7, value = "CD103")
#switch to RNA assay
DefaultAssay(Clus7) <- "RNA"
```

Look at differences of genes of interest by violin plot (between CD103+ vs CD103- cells) for Supplemental Fig. S8C.
```{r}
comparisons <- list(c("Pos", "Neg"))
VlnPlot(object = Clus7, features = c('ITGAE', "PDCD1", "LAG3", "TOX", "KLF2", "TCF7", "SELL", "IL7R"), pt.size = F, cols = c("#DC0000B2","#6A6767")) + stat_compare_means(comparisons = comparisons, method = "wilcox.test", label = "p.format")
```

Look at differences of proteins of interest by violin plot (between CD103+ vs CD103- cells) for Supplemental Fig. S8D.
```{r}
DefaultAssay(Clus7) <- "integratedADT"
VlnPlot(object = Clus7, features = c(".CD4", ".CD8a", ".CD69", ".CD279", ".CD127", ".KLRG1"), pt.size = F, cols = c("#DC0000B2","#6A6767"))
+ stat_compare_means(comparisons = list(c("Pos", "Neg")) ,method = "wilcox.test", label = "p.format")

```

