---
title: "Internal scRNAseq analysis of T cells"
author: "Gillian Savage"
date: "1/6/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(ggplot2)
library(patchwork)
library(reshape2)
library(RColorBrewer)
library(dplyr)
library(dittoSeq)
library(tidyverse)
library(ggsci)
library(ggsignif)
library(enrichplot)
library(presto)
library(lubridate)
library(pheatmap)
library(ggpubr)
library(hues)
library(igraph)
library(cluster)
library(fgsea)
```

#Load data

Load seurat object of T cells:
```{r}
DLBCL_Tcells <- readRDS("~/trms_dlbcl/internal_data/all_T_clustered.rds")
```

Add metadata to seurat object
```{r}
#load metadata
md <- read.csv("~/trms_dlbcl/internal_data/metadata.csv", stringsAsFactors = FALSE)

#create new df joining md to seurat samples based on original sample ID
add_data <- left_join(DLBCL_Tcells[["orig.ident"]], md)

#covert df to have rownames of cell ids
row.names(add_data) <- row.names(DLBCL_Tcells[[]])

#finally add metadata to seurat object
DLBCL_Tcells <- AddMetaData(DLBCL_Tcells, metadata = add_data)

```

## T cell cluster analysis (related to Figure 2)

All T cell clusters (Figure 2A)
```{r}
DimPlot(DLBCL_Tcells, group.by = "seurat_clusters", cols = c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#EE4C97", "#6A6767","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"), label = TRUE)
```

For labeling of T cell clusters we used a custom list of genes related to well known T cell phenotypes and generated a heatmap to visualize expression of these genes (for generation of Figure 2B)
```{r}
# make a empty data table
corrected.cluster.means <- data.table()
```

```{r}

# seurat_clusters is the column where you store the cluster information
# this part will calculate the average expression of genes in each cell cluster
for (i.clust in levels(DLBCL_Tcells@meta.data$seurat_clusters)) {
	i.clust.means <- data.frame(rowMeans(DLBCL_Tcells[["RNA"]]$data[,which(DLBCL_Tcells@meta.data$seurat_clusters== i.clust)]))
	colnames(i.clust.means) <- i.clust
	rownames(i.clust.means) <- rownames(DLBCL_Tcells[["RNA"]]$data)
	if (nrow(corrected.cluster.means) < 1) {
		corrected.cluster.means <- i.clust.means
	}
	else {
		corrected.cluster.means <- cbind(corrected.cluster.means, i.clust.means)
	}
}
```

```{r}

# read the markers list
comp.func.data <- read.delim("~/trms_dlbcl/internal_data/custom_T_sig_fig.txt", stringsAsFactors = FALSE)

comp.func.data$Component <- as.factor(comp.func.data$Component)
```

```{r}
# gene annotation 
comp.func.ann <- comp.func.data[,"Component", drop = FALSE]
rownames(comp.func.ann) <- comp.func.data$Gene
# assign colors for the genes
comp.func.ann.colors <- list(Component = iwanthue(nlevels(comp.func.data$Component)))
names(comp.func.ann.colors$Component) <- levels(comp.func.data$Component)
comp.func.rownames <- apply(comp.func.data, 1, function(x) {if (x["Alt_name"] == "") { return(x["Gene"]) } else { return(paste0(x["Gene"], " (", x["Alt_name"], ")")) }})

# gaps in the heatmap
gaps.row <- which(!duplicated(comp.func.ann$Component))[-1] - 1

# subset the matrix to keep only the markers
corrected.comp.mat <- as.matrix(corrected.cluster.means[comp.func.data$Gene,])


```

Plot heatmap (Figure 2B)
```{r}
(pheatmap(corrected.comp.mat, 
          scale = "row", 
          cluster_rows = FALSE, 
          cluster_cols = TRUE, 
          labels_row = comp.func.rownames,
					annotation_colors = comp.func.ann.colors, 
					annotation_row = comp.func.ann, 
					annotation_names_row = FALSE,
					gaps_row = gaps.row,
          color = colorRampPalette(c("navy", "white", "red"))(50)))

```

Overlay of whether T cells came from rLN or lymphoma patient (Figure 2C)
```{r}
DimPlot(DLBCL_Tcells, group.by = "Control_or_lymphoma", cols = c("#BC3C29FF","#6F99ADFF"))
```

# Analysis of CD103+ T cells (related to Figure 2)

The gene for CD103 is ITGAE so first we identified cells that expressed ITGAE using an expression cutoff off value of 0.5
```{r}
#first switch active assay to RNA
DefaultAssay(DLBCL_Tcells) <- "RNA"

#evaluate expression in each cluster
DLBCL_Tcells <- SetIdent(DLBCL_Tcells, value = "seurat_clusters")

#volcano plot of ITGAE expression across clusters (Figure 2E)
VlnPlot(object = DLBCL_Tcells, features = 'ITGAE', cols = c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#EE4C97", "#6A6767","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"))+ geom_hline(yintercept = 0.5) # add line at 0.5 to evaluate this cutoff value
```

Expression value cutoff of 0.5 seems appropriate based on violin plot
Add to seurat column of cells positive for CD103
```{r}
# create new column in the metadata for Neg/Pos status of ITGAE/CD103
DLBCL_Tcells@meta.data$CD103 <- 'Neg'
DLBCL_Tcells@meta.data$CD103[which(DLBCL_Tcells@assays$RNA$counts["ITGAE",]>0.5)] <- 'Pos'
```

Plot on UMAP which cells are positive for ITGAE (Fig 2D)
```{r}
#Set indent to CD103
DLBCL_Tcells <- SetIdent(DLBCL_Tcells, value = "CD103")

DimPlot(DLBCL_Tcells, label = F, cols = c("Neg" = "#C5C6C7", "Pos" = "#BC3C29"))

```

Look at proportion of T cells in each patient are positive for ITGAE between rLN and lymphoma patients (Fig 2F)
```{r}
dittoFreqPlot(DLBCL_Tcells, "CD103",
    sample.by = "orig.ident", group.by = "Control_or_lymphoma", vars.use = "Pos",color.panel = c("#DC0000B2", "#7fb8da")) +
  geom_signif(comparisons = list(c("RLN", "LBCL")),
              step_increase = 0.2, y_position = 0.45,
              test = t.test,textsize = 3.5) 

```

Now we will look just at cells positive for ITGAE and recluster/map just these T cells.
```{r}
set.seed(123)
#make sure ident is set to CD103 ident
DLBCL_Tcells <- SetIdent(DLBCL_Tcells, value = "CD103")

#subset out just T cells positive for ITGAE/CD103
CD103Tcells <- subset(x=DLBCL_Tcells, idents = c("Pos"))

```

PCA and dimensionality reduction of ITGAE+ cells:

```{r}
#make sure active assay is SCT
DefaultAssay(CD103Tcells) <- "SCT"


CD103Tcells <- RunPCA(CD103Tcells, verbose = F)
CD103Tcells <- RunUMAP(CD103Tcells, dims = 1:10, verbose = F)
CD103Tcells <- FindNeighbors(CD103Tcells, dims = 1:10, k.param = 12, verbose = F)
CD103Tcells <- FindClusters(CD103Tcells, verbose = F)
table(CD103Tcells[[]]$seurat_clusters)

#view UMAP with clusters
set.seed(123)
DimPlot(CD103Tcells, label = T)
```

**Determining optimal number of clusters for ITGAE+ T cells**

First try k-values of 10, 15, 20. 30
```{r}
#k=10
    CD103Tcells <- FindNeighbors(CD103Tcells, k.param = 10)
    CD103Tcells<-FindClusters(CD103Tcells, resolution=0.8)
    CD103Tcells@meta.data$SCT_res0.8_k10 <- CD103Tcells@meta.data$SCT_snn_res.0.8
    table(CD103Tcells@meta.data$seurat_clusters)
    
#k=15
    CD103Tcells <- FindNeighbors(CD103Tcells, k.param = 15)
    CD103Tcells<-FindClusters(CD103Tcells, resolution=0.8)
    CD103Tcells@meta.data$SCT_res0.8_k15 <- CD103Tcells@meta.data$SCT_snn_res.0.8
    table(CD103Tcells@meta.data$seurat_clusters)

#k=20
    CD103Tcells <- FindNeighbors(CD103Tcells, k.param = 20)
    CD103Tcells<-FindClusters(CD103Tcells, resolution=0.8)
    CD103Tcells@meta.data$SCT_res0.8_k20 <- CD103Tcells@meta.data$SCT_snn_res.0.8
    table(CD103Tcells@meta.data$seurat_clusters)
    
    
#k=30
    CD103Tcells <- FindNeighbors(CD103Tcells, k.param = 30)
    CD103Tcells<-FindClusters(CD103Tcells, resolution=0.8)
    CD103Tcells@meta.data$SCT_res0.8_k30 <- CD103Tcells@meta.data$SCT_snn_res.0.8
    table(CD103Tcells@meta.data$seurat_clusters)


```

Calculate Siloutte Index for each k-value
```{r}
clusterings <- grepl("SCT_res0.8",colnames(CD103Tcells@meta.data))
dist.matrix <- dist(x = Embeddings(object = CD103Tcells[["pca"]])[, 1:10])
clusterings <- colnames(CD103Tcells@meta.data)[clusterings]
for (c in clusterings) {
  require(cluster) # contains silhouette function.
  sil <- silhouette(as.numeric(as.character(CD103Tcells@meta.data[,c])), dist=dist.matrix)
   print(paste(c))
  print(mean(sil[,3]))
  print(head(sil, n=20))
}
```

Decided to go ahead with k = 20 so now will try different resolutions to fine tune cluster number
```{r}
CD103Tcells <- FindNeighbors(CD103Tcells, k.param = 20, dims = 1:10)

for(res in c(0.2, 0.5, 0.8, 1, 1.5)) {
    CD103Tcells<-FindClusters(CD103Tcells, resolution=res)
    print(paste(res))
    print(table(CD103Tcells@meta.data$seurat_clusters))
}

```

Calculate Siloutte Index for each resolution:
```{r}
clusterings <- grepl("SCT_snn_res",colnames(CD103Tcells@meta.data))
dist.matrix <- dist(x = Embeddings(object = CD103Tcells[["pca"]])[, 1:15])
clusterings <- colnames(CD103Tcells@meta.data)[clusterings]
for (c in clusterings) {
  require(cluster) # contains silhouette function.
  sil <- silhouette(as.numeric(as.character(CD103Tcells@meta.data[,c])), dist=dist.matrix)
   print(paste(c))
  print(mean(sil[,3]))
  print(head(sil, n=20))
}
```

Calculate Rand Index for each resolution:
```{r}

for (c in clusterings) {
  for (c2 in clusterings) {
    ARI <- compare(CD103Tcells@meta.data[,c], CD103Tcells@meta.data[,c2], method="adjusted.rand")
    print(paste(c, c2, ARI))
  }
}

```

0.5 and 0.8 and 0.2 and 1 look quite close so decided between those using number of marker genes per cluster
```{r}
#prep for find markers
CD103Tcells <- PrepSCTFindMarkers(object = CD103Tcells, assay = "SCT", verbose = TRUE)
for (c in clusterings) {
  markers <- FindAllMarkers(CD103Tcells, group.by=c, logfc.threshold = -Inf, only.pos=TRUE, max.cells.per.ident=100, verbose=FALSE)
  markers <- markers[ markers[,"p_val_adj"] < 0.05, ]
  print(paste(c))
  print(table(markers$cluster))
}

```

Resolution of 0.2 looks best. Re-cluster with k=20 and resolution = 0.2
```{r}
CD103Tcells <- FindNeighbors(CD103Tcells, k.param = 20, dims = 1:10, verbose = F)
CD103Tcells<-FindClusters(CD103Tcells, resolution=0.2)

```

Plot UMAP of ITGAE+ T cell clusters (Fig 2G)
```{r}
CD103Tcells <- SetIdent(CD103Tcells, value="seurat_clusters")
DimPlot(CD103Tcells,label = T, cols = c("#BC3C29","#E18727", "#0072B5", "#EE4C97", "#7876B1", "#6F99AD", "#FFDC91"))
```

Save just the CD103+ T cells seurat file for downstream analysis
```{r}
saveRDS(CD103Tcells, file = "~/trms_dlbcl/seurat_CD103Tcells.rds")
```

Using Custom T cell markers/signatures based on literature/top genes to make heatmap for Supp fig S5A
```{r}
# make a empty data table
corrected.cluster.means <- data.table()
```

```{r}

# seurat_clusters is the column where you store the cluster information
# this part will calculate the average expression of genes in each cell cluster
for (i.clust in levels(CD103Tcells@meta.data$seurat_clusters)) {
	i.clust.means <- data.frame(rowMeans(CD103Tcells[["RNA"]]$data[,which(CD103Tcells@meta.data$seurat_clusters== i.clust)]))
	colnames(i.clust.means) <- i.clust
	rownames(i.clust.means) <- rownames(CD103Tcells[["RNA"]]$data)
	if (nrow(corrected.cluster.means) < 1) {
		corrected.cluster.means <- i.clust.means
	}
	else {
		corrected.cluster.means <- cbind(corrected.cluster.means, i.clust.means)
	}
}
```

```{r}
# gene annotation 
comp.func.ann <- comp.func.data[,"Component", drop = FALSE]
rownames(comp.func.ann) <- comp.func.data$Gene
# assign colors for the genes
comp.func.ann.colors <- list(Component = iwanthue(nlevels(comp.func.data$Component)))
names(comp.func.ann.colors$Component) <- levels(comp.func.data$Component)
comp.func.rownames <- apply(comp.func.data, 1, function(x) {if (x["Alt_name"] == "") { return(x["Gene"]) } else { return(paste0(x["Gene"], " (", x["Alt_name"], ")")) }})

# gaps in the heatmap
gaps.row <- which(!duplicated(comp.func.ann$Component))[-1] - 1

# subset the matrix to keep only the markers
corrected.comp.mat <- as.matrix(corrected.cluster.means[comp.func.data$Gene,])


```

Plot heatmap (Supplementary Fig S5A)
```{r}
(pheatmap(corrected.comp.mat, 
          scale = "row", 
          cluster_rows = FALSE, 
          cluster_cols = TRUE, 
          labels_row = comp.func.rownames,
					annotation_colors = comp.func.ann.colors, 
					annotation_row = comp.func.ann, 
					annotation_names_row = FALSE,
					gaps_row = gaps.row,
          color = colorRampPalette(c("navy", "white", "red"))(50)))

```

Overlay of rLN or DLBCL samples over UMAP of ITGAE+ T cells (Supplementary Fig S5B)
```{r}
DimPlot(CD103Tcells, group.by = "Control_or_lymphoma", cols = c("#BC3C29FF","#6F99ADFF"))
```

Now add a new column of general T cell phenotypes determined based on expression of genes in heatmap of Supplementary Figure S5A
```{r}
#get number of clusters
nclusters <- length(unique(CD103Tcells@meta.data$seurat_clusters))
# Create a vector where the names of the vector are the cluster number
# and the values of the vector are the annotated names.
cluster_annotation2 <- rep("unannotated", nclusters)
names(cluster_annotation2) <- levels(CD103Tcells@meta.data$seurat_clusters)
cluster_annotation2["0"] = "Activated/Effector"
cluster_annotation2["1"] = "Activated/Effector"
cluster_annotation2["2"] = "Memory/Naive"
cluster_annotation2["3"] = "Tfh"
cluster_annotation2["4"] = "Treg"
cluster_annotation2["5"] = "Proliferating"
cluster_annotation2["6"] = "Tfh"


CD103Tcells <- SetIdent(CD103Tcells, value=CD103Tcells@meta.data$seurat_clusters)
CD103Tcells <- RenameIdents(CD103Tcells, cluster_annotation2)
CD103Tcells@meta.data$cellID <- Idents(CD103Tcells)
head(CD103Tcells@meta.data)
```

Create pie chart of clusters (Fig 2G)
```{r}
#get number of cells per cluster
prop <- data.frame(table(CD103Tcells[[]]$cellID))

#create piechart
ggplot(prop, aes(x="", y=Freq, fill=Var1)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() +
  labs(fill = "Cluster") +
  geom_text(aes(label = Freq),
            position = position_stack(vjust = 0.5)) + scale_fill_manual(values=c("#BC3C29","#0072B5", "#EE4C97", "#7876B1", "#6F99AD", "#FFDC91"))
```

Look at proportion of ITGAE+ T cells in each cluster between rLN and DLBCL patients (Fig 2H)
```{r}
dittoFreqPlot(CD103Tcells, "seurat_clusters",
   sample.by = "orig.ident", group.by = "Control_or_lymphoma",color.panel = c("#DC0000B2", "#7fb8da"), split.nrow = 1)
```

*Calculation of enrichment Score of CD103+ T cells in DLBCL vs rLN patients for proliferation, cytotoxicity, exhaustion and activation gene sets*
Load gene sets:
```{r}
#load genes to be used for proliferation scoring

proliferation.genes <- c("ZWINT", "E2F1", "FEN1", "FOXM1", "H2AFZ", "HMGB2", "MCM2", "MCM3", "MCM4", "MCM5", "MCM6", "MKI67", "MYBL2", "PCNA", "CCND1", "AURKA", "BUB1", "TOP2A", "TYMS", "DEK", "CCNB1")

#load genes to be used for cytoxicity scoring

cytotox.genes <- c("KLRF1", "GNLY", "CTSW", "NKG7", "KLRD1", "GZMA", "ADGRG1", "CST7", "KLRK1", "FASLG", "HCST", "KLRB1", "ITGB1", "GZMB", "PRF1")

#load genes to be used for exhaustion scoring

exhaustion.genes <- c("HAVCR2", "LAG3", "TIGIT", "CTLA4", "PDCD1", "LAYN", "BTLA", "CD160")

#load genes to be used for activation scoring

activation.genes <- c("IL26", "IL13", "PARP8", "CD63", "IFNG", "GZMB", "STAT1", "TNF", "HLA-DRA", "HLA-DRB5", "HLA-DRB1")

```

Enrichment score calculations:
```{r}
#add score for each cell for proliferation
CD103Tcells <- AddModuleScore(CD103Tcells,
                  features = list(proliferation.genes),
                  name="Prolif.score")

#add score for each cell for cytoxicity
CD103Tcells <- AddModuleScore(CD103Tcells,
                  features = list(cytotox.genes),
                  name="Cytotox.score")

#add score for each cell for exhaustion
CD103Tcells <- AddModuleScore(CD103Tcells,
                  features = list(exhaustion.genes),
                  name="EX.score")

#add score for each cell for activation
CD103Tcells <- AddModuleScore(CD103Tcells,
                  features = list(activation.genes),
                  name="Activation.score")


#pull out metatdata
CD103.df <- data.frame(CD103Tcells@meta.data)

#subset for scores
CD103.df.scores <- CD103.df[, c("seurat_clusters", "Control_or_lymphoma", "Prolif.score1", "Cytotox.score1", "EX.score1", "Activation.score1")]

head(CD103.df.scores)

```

Generate violin plots of scores (Fig 2I)
```{r}
#proliferation
ggplot(CD103.df.scores,aes(x=Control_or_lymphoma,y=Prolif.score1,fill=Control_or_lymphoma))+geom_violin()+theme_classic()+ggtitle("Proliferation score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#BC3C29FF","#6F99ADFF"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-0.5, 1.8)+
  geom_signif(comparisons = list(c("RLN", "LBCL")),
              step_increase = 0.2,
              test = wilcox.test,textsize = 3.5, test.args=list(alternative = "two.sided", var.equal = FALSE, paired=FALSE)) 

#cytotoxicity
ggplot(CD103.df.scores,aes(x=Control_or_lymphoma,y=Cytotox.score1,fill=Control_or_lymphoma))+geom_violin()+theme_classic()+ggtitle("Cytotoxicity score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#BC3C29FF","#6F99ADFF"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-0.5, 1.8)+
  geom_signif(comparisons = list(c("RLN", "LBCL")),
              step_increase = 0.2,
             # map_signif_level = T,
              test = wilcox.test,textsize = 3.5, test.args=list(alternative = "two.sided", var.equal = FALSE, paired=FALSE))

#exhaustion
ggplot(CD103.df.scores,aes(x=Control_or_lymphoma,y=EX.score1,fill=Control_or_lymphoma))+geom_violin()+theme_classic()+ggtitle("Exhaution score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#BC3C29FF","#6F99ADFF"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-0.5, 1.8) +
  geom_signif(comparisons = list(c("RLN", "LBCL")),
              step_increase = 0.2,
              #map_signif_level = T,
              test = wilcox.test,textsize = 3.5, test.args=list(alternative = "two.sided", var.equal = FALSE, paired=FALSE))

#activation
ggplot(CD103.df.scores,aes(x=Control_or_lymphoma,y=Activation.score1,fill=Control_or_lymphoma))+geom_violin()+theme_classic()+ggtitle("Activation score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#BC3C29FF","#6F99ADFF"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-0.5, 1.8)+
  geom_signif(comparisons = list(c("RLN", "LBCL")),
              step_increase = 0.2,
             # map_signif_level = T,
              test = wilcox.test,textsize = 3.5,  test.args=list(alternative = "two.sided", var.equal = FALSE, paired=FALSE))

```

#Generate plots looking at TCR and TGFb signaling between CD103+ and CD103- T cells (Figure 2J-K)
```{r}
#load pathways
TCR.sig <- getGmt("~/trms_dlbcl/internal_data/KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY.gmt")
TGF.sig <- getGmt("~/trms_dlbcl/internal_data/KEGG_TGF_BETA_SIGNALING_PATHWAY.gmt")

# Extract gene sets into a named list
tcr_list <- lapply(TCR.sig, geneIds)
names(tcr_list) <- sapply(TCR.sig, function(x) x@setName)

tgf_list <- lapply(TGF.sig, geneIds)
names(tgf_list) <- sapply(TGF.sig, function(x) x@setName)

TCR.sig <- list(tcr_list[["KEGG_T_CELL_RECEPTOR_SIGNALING_PATHWAY"]])
TGF.sig <- list(tgf_list[["KEGG_TGF_BETA_SIGNALING_PATHWAY"]])

```
Enrichment score calculations:
```{r}
DefaultAssay(DLBCL_Tcells) <- "RNA"
#add score for each cell for TCR and TGFb signaling
DLBCL_Tcells <- AddModuleScore(DLBCL_Tcells,
                  features = TCR.sig,
                  name="TCR.score")

DLBCL_Tcells <- AddModuleScore(DLBCL_Tcells,
                  features = TGF.sig,
                  name="TGF.score")

```
```{r}
#pull out metatdata
TCRsig.df <- data.frame(DLBCL_Tcells@meta.data)

#subset for scores
TCRsig.df <- TCRsig.df[, c("seurat_clusters", "CD103", "TCR.score1", "TGF.score1", "Control_or_lymphoma")]

head(TCRsig.df)

```
Plot enrichment scores, separating by cells that come from LBCL vs rLN patients
```{r}
#plot TCRsig score
ggplot(TCRsig.df,aes(x=CD103,y=TCR.score1,fill=CD103))+geom_violin(position = "dodge")+theme_classic()+ggtitle("TCR signaling pathway score")+scale_x_discrete("")+
  facet_grid(. ~ Control_or_lymphoma) +
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#6A6767", "#DC0000B2"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.2) + 
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pos", "Neg")),
                      hide.ns = TRUE, label.y = 0.6, size = 3)
#plot TGFb sig score
ggplot(TCRsig.df,aes(x=CD103,y=TGF.score1,fill=CD103))+geom_violin(position = "dodge")+theme_classic()+ggtitle("TGF signaling pathway score")+scale_x_discrete("")+
  facet_grid(. ~ Control_or_lymphoma) +
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#6A6767", "#DC0000B2"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.2) + 
  stat_compare_means(label = "p.format", method = "wilcox.test", comparisons = list(c("Pos", "Neg")),
                      hide.ns = TRUE, label.y = 0.4, size = 3)

```

# Looking for a TRM-like cluster in DLBCL

Look at Trm score within each cluster by enrichment scoring (Fig 3A)
```{r}
#load genes associated with Trm sig
Trm.genes <- c("ITGAE",	"AHR",	"ANXA1",	"CTLA4",	"CXCR6",	"DUSP5",	"EPAS1",	"GZMB",	"IFNG",	"ITGAV",	"KLRC1",	"LMNA",	"RBPJ",	"TNFRSF9",	"XCL1",	"PDCD1",	"DUSP6",	"IL10",	"IL2",	"KCNK5",	"CXCL13",	"ITGA1",	"RGS1", "PRDM1", "RUNX3", "CD69", "DAPK2")

DefaultAssay(DLBCL_Tcells) <- "SCT"
#add score for each cell for tissue residency
DLBCL_Tcells <- AddModuleScore(DLBCL_Tcells,
                  features = list(Trm.genes),
                  name="Trm.score")

#pull out metatdata
allT.df <- data.frame(DLBCL_Tcells@meta.data)

#plot on UMAP
FeaturePlot(DLBCL_Tcells,
            features = "Trm.score1", label = F, repel = TRUE, alpha = 4) +
            scale_colour_gradientn(colours = brewer.pal(n = 9, name = "OrRd"))
```



Plot gene expression of core TRM associated genes or genes associated with circulation (S1PR1, CCR7, SELL) on UMAP of all T cells (Fig 3B)
```{r}
FeaturePlot(DLBCL_Tcells, features = c("ITGAE", "PDCD1", "ENTPD1", "GZMB", "PRDM1", "CD69", "LAG3", "CXCR6", "RUNX3", "S1PR1", "CCR7", "SELL"), cols = c("lightgrey", "navy"))
```

Generate heatmap of Trm associated genes (Fig 3C)
```{r}
# read the markers list
comp.func.data <- read.delim("~/trms_dlbcl/internal_data/TRM_heatmap.txt", stringsAsFactors = FALSE)

comp.func.data$Component <- as.factor(comp.func.data$Component)

# gene annotation 
comp.func.ann <- comp.func.data[,"Component", drop = FALSE]
rownames(comp.func.ann) <- comp.func.data$Gene
# assign colors for the genes
comp.func.ann.colors <- list(Component = iwanthue(nlevels(comp.func.data$Component)))
names(comp.func.ann.colors$Component) <- levels(comp.func.data$Component)
comp.func.rownames <- apply(comp.func.data, 1, function(x) {if (x["Alt_name"] == "") { return(x["Gene"]) } else { return(paste0(x["Gene"], " (", x["Alt_name"], ")")) }})

# gaps in the heatmap
gaps.row <- which(!duplicated(comp.func.ann$Component))[-1] - 1

# subset the matrix to keep only the markers
corrected.comp.mat <- as.matrix(corrected.cluster.means[comp.func.data$Gene,])

(pheatmap(corrected.comp.mat, 
          scale = "row", 
          cluster_rows = FALSE, 
          cluster_cols = TRUE, 
          labels_row = comp.func.rownames,
					annotation_colors = comp.func.ann.colors, 
					annotation_row = comp.func.ann, 
					annotation_names_row = FALSE,
					gaps_row = gaps.row,
          color = colorRampPalette(c("navy", "white", "red"))(50)))

```

Cluster 9 has highest Trm score. Perform gene set enrichment analysis using previously published gene signatures

load custom Trm gene sets:
```{r}
#load and read gmt file
gmt.file <- "~/trms_dlbcl/internal_data/Trm_genesets.gmt"
Trm.sigs <- gmtPathways(gmt.file)
str(head(Trm.sigs))

```

First perform wilcox rank sum test to get rank of all genes for each cluster
```{r}
#first switch active assay to RNA
DefaultAssay(DLBCL_Tcells) <- "RNA"

#then wilcox test with presto
allT.genes <- wilcoxauc(DLBCL_Tcells, 'seurat_clusters')

head(allT.genes)

# we have all the genes for each cluster
dplyr::count(allT.genes, group)
```

```{r}
# select only the feature and auc columns for fgsea, which statistics to use is an open question
C9.genes<- allT.genes %>%
  dplyr::filter(group == "9") %>%
  arrange(desc(auc)) %>% 
  dplyr::select(feature, auc)


ranks_C9<- deframe(C9.genes)

head(ranks_C9)
```

Perform GSEA for Trm signatures:
```{r}
fgseaRes<- fgsea(Trm.sigs, stats = ranks_C9, nperm = 2000)

#tidy up data
fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))


fgseaResTidy %>% 
  dplyr::select(-leadingEdge, -ES, -nMoreExtreme) %>% 
  arrange(padj) %>% 
  head()
```

```{r}
fgseaRes <- fgseaMultilevel(
  Trm.sigs,
  stats = ranks_C9,
  minSize = 1,
  maxSize = Inf,
  eps = 1e-10,
  scoreType = c("std", "pos", "neg"),
  nproc = 0,
  gseaParam = 1,
  BPPARAM = NULL,
  nPermSimple = 2000,
  absEps = NULL
)

#tidy up data
fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  arrange(desc(NES))


fgseaResTidy %>% 
  dplyr::select(-leadingEdge, -ES) %>% 
  arrange(padj) %>% 
  head()
```

Generate lollipop plot of enrichment scores and p-values (Fig 3D)
```{r}
#pathways of interest
paths <- c("Savas_et_al_Trm_UP", "Molodtsov_et_al_LN_Trm",  "Kumar_TRM_coresig","Szabo_CD8_Trm_rest", "Szabo_CD4_Trm_rest", "Szabo_CD4_Trm_act", "Hasen_et_al_Trm_UP", "Milner_et_al_Trm")
#subset just those pathways
fgsea.sub <- fgseaResTidy %>% subset(pathway %in% paths)

#order and pathway to factor
fgsea.sub <- fgsea.sub[order(fgsea.sub$NES, decreasing = F),]
fgsea.sub$pathway <- factor(fgsea.sub$pathway, levels = c(fgsea.sub$pathway))

# Create the plot:
ggplot(fgsea.sub, aes(x = pathway, y = NES)) +
  geom_segment(aes(x = pathway, xend = pathway, y = 0, yend = NES, fill = -log(padj)), lwd = 1) +
  geom_point(pch = 19, aes(size=size, color = -log(padj))) + scale_color_gradient(low = "blue", high = "red")+
  theme_bw(base_size=10, base_family = "Helvetica") +
  theme(axis.text=element_text(size=12, colour = "black"),
        axis.title=element_text(size=14)) +
  ggtitle("TRM pathways") +
  coord_flip()

```

Perform enrichment scoring for proliferation, cytotox, exhaustion, and activation gene sets on T cell clusters that are characterized as activated/effector (clusters 1, 6, 7, 9, 11)

```{r}
#add score for each cell for proliferation
DLBCL_Tcells <- AddModuleScore(DLBCL_Tcells,
                  features = list(proliferation.genes),
                  name="Prolif.score")

#add score for each cell for cytoxicity
DLBCL_Tcells <- AddModuleScore(DLBCL_Tcells,
                  features = list(cytotox.genes),
                  name="Cytotox.score")

#add score for each cell for exhaustion
DLBCL_Tcells <- AddModuleScore(DLBCL_Tcells,
                  features = list(exhaustion.genes),
                  name="EX.score")

#add score for each cell for activation
DLBCL_Tcells <- AddModuleScore(DLBCL_Tcells,
                  features = list(activation.genes),
                  name="Activation.score")

#pull out metatdata
all.Tcells.df <- data.frame(DLBCL_Tcells@meta.data)

#subset for scores
all.Tcells.df.scores <- all.Tcells.df[, c("seurat_clusters", "CD103", "Prolif.score1", "Cytotox.score1", "EX.score1", "Activation.score1")]

head(all.Tcells.df.scores)

```

Subset just effector columns which are clusters 1, 6, 7, 9, 11 to compare these scores between just effector T cell clusters
```{r}
eff.Tcell.scores <- all.Tcells.df.scores[all.Tcells.df.scores$seurat_clusters == c('1', "6", "7", "9", "11"),]
head(eff.Tcell.scores)
```

Generate violin plots for each score within each cluster (Fig 3E)
```{r}
#proliferation
ggplot(eff.Tcell.scores,aes(x=seurat_clusters,y=Prolif.score1,fill=seurat_clusters))+geom_violin()+theme_classic()+ggtitle("Proliferation score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#0072B5", "#ffc445", "#EE4C97","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-0.5, 2.5) + 
  stat_compare_means(method = "anova", label.y = 2)+        # Add global annova p-value
  stat_compare_means(label = "p.format", method = "wilcox.test",
                     ref.group = ".all.", hide.ns = TRUE, size = 3)  


#cytotoxicity
ggplot(eff.Tcell.scores,aes(x=seurat_clusters,y=Cytotox.score1,fill=seurat_clusters))+geom_violin()+theme_classic()+ggtitle("Cytotoxicity score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#EE4C97", "#6A6767","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-0.5, 2.5)+ 
  stat_compare_means(method = "anova", label.y = 2)+        # Add global annova p-value
  stat_compare_means(label = "p.format", method = "wilcox.test",
                     ref.group = ".all.", hide.ns = TRUE, size = 3)

#exhaustion
 ggplot(eff.Tcell.scores,aes(x=seurat_clusters,y=EX.score1,fill=seurat_clusters))+geom_violin()+theme_classic()+ggtitle("Exhaustion score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#EE4C97", "#6A6767","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-0.5, 2.5)+ 
  stat_compare_means(method = "anova", label.y = 2)+        # Add global annova p-value
  stat_compare_means(label = "p.format", method = "wilcox.test",
                     ref.group = ".all.", hide.ns = TRUE, size = 3)

#activation
ggplot(eff.Tcell.scores,aes(x=seurat_clusters,y=Activation.score1,fill=seurat_clusters))+geom_violin()+theme_classic()+ggtitle("Activation score")+scale_x_discrete("")+
  theme(title = element_text(face = "plain",size = 12),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        aspect.ratio = 1.2,
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank(),
        plot.title=element_text(hjust = 0.5))+ scale_fill_manual(values=c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#EE4C97", "#6A6767","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"))+
  geom_boxplot(width=0.2,position=position_dodge(0.9), outlier.colour = "black", fill="white", outlier.size=0.5)+ylim(-0.5, 2.5)+ 
  stat_compare_means(method = "anova", label.y = 2)+        # Add global annova p-value
  stat_compare_means(label = "p.format", method = "wilcox.test",
                     ref.group = ".all.", hide.ns = TRUE, size = 3)

```

Graph the proportion of T cells in cluster 9 (TRM cluster) per DLBCL vs rLN patient (Fig 3F)
```{r}
dittoFreqPlot(DLBCL_Tcells, "seurat_clusters",
    sample.by = "orig.ident", group.by = "Control_or_lymphoma", vars.use = "9",color.panel = c("#DC0000B2", "#7fb8da")) +
  geom_signif(comparisons = list(c("RLN", "LBCL")),
              step_increase = 0.2, y_position = 0.75,
              test = t.test,textsize = 3.5) 
```
