---
title: "Internal scRNAseq Preprocessing & Integration"
author: "Gillian Savage"
date: "1/6/2025"
output: html_document
---

Load Libraries
```{r}
#Load Libraries

library(Seurat)
library(patchwork)
library(reshape2)
library(RColorBrewer)
library(dplyr)

require(SingleR)
require(SingleCellExperiment)
require(celldex)
```

## Load Data and generate Seurat Object

Have DLBCL or HGL samples and rLN samples that have been processed by CellRanger to generate counts matrices. Count matrices are loaded individually before being combined. Counts files are organized by patient sample, with each having its own directory.

```{r}
setwd("~/trms_dlbcl")
#load all counts matrix outputs from Cellranger
RLN02 <- Read10X(data.dir = "./counts_files/RLN02/")
RLN03 <- Read10X(data.dir = "./counts_files/RLN03/")
RLN04 <- Read10X(data.dir = "./counts_files/RLN04/")
RLN05 <- Read10X(data.dir = "./counts_files/RLN05/")
CCS0025 <- Read10X(data.dir = "./counts_files/CCS0025/")
GE0539 <- Read10X(data.dir = "./counts_files/GE0539/")
DLC0087 <- Read10X(data.dir = "./counts_files/DLC0087/")
DLC0341 <- Read10X(data.dir = "./counts_files/DLC0341/")
GE0287 <- Read10X(data.dir = "./counts_files/GE0287/")
DLC0169 <- Read10X(data.dir = "./counts_files/DLC0169/")
CLC00277 <- Read10X(data.dir = "./counts_files/CLC00277/")
GE0505 <- Read10X(data.dir = "./counts_files/GE0505/")
CLC00275 <- Read10X(data.dir = "./counts_files/CLC00275/")
CCS0099 <- Read10X(data.dir = "./counts_files/CCS0099/")
CCS0277 <- Read10X(data.dir = "./counts_files/CCS0277/")
CCS2010 <- Read10X(data.dir = "./counts_files/CCS2010/")

```

Each count matrix is then converted into a seurat object

```{r}
#create seurat objects for each
RLN02_seur <- CreateSeuratObject(RLN02, project = "RLN02")
RLN03_seur <- CreateSeuratObject(RLN03, project = "RLN03")
RLN04_seur <- CreateSeuratObject(RLN04, project = "RLN04")
RLN05_seur <- CreateSeuratObject(RLN05, project = "RLN05")
CCS0025_seur <- CreateSeuratObject(CCS0025, project = "CCS0025")
GE0539_seur <- CreateSeuratObject(GE0539, project = "GE0539")
DLC0087_seur <- CreateSeuratObject(DLC0087, project = "DLC0087")
DLC0341_seur <- CreateSeuratObject(DLC0341, project = "DLC0341")
GE0287_seur <- CreateSeuratObject(GE0287, project = "GE0287")
DLC0169_seur <- CreateSeuratObject(DLC0169, project = "DLC0169")
CLC00277_seur <- CreateSeuratObject(CLC00277, project = "CLC00277")
GE0505_seur <- CreateSeuratObject(GE0505, project = "GE0505")
CLC00275_seur <- CreateSeuratObject(CLC00275, project = "CLC00275")
CCS0099_seur <- CreateSeuratObject(CCS0099, project = "CCS0099")
CCS0277_seur <- CreateSeuratObject(CCS0277, project = "CCS0277")
CCS2010_seur <- CreateSeuratObject(CCS2010, project = "CCS2010")


#remove matrices to save space since they are no longer needed
rm(RLN02)
rm(RLN03)
rm(RLN04)
rm(RLN05)
rm(CCS0025)
rm(GE0539)
rm(DLC0087)
rm(DLC0341)
rm(GE0287)
rm(DLC0169)
rm(CLC00277)
rm(GE0505)
rm(CLC00275)
rm(CCS0099)
rm(CCS0277)
rm(CCS2010)
```

## QC & Sample Filtering

For each sample we want to remove dead or damaged cells that may interfere with our results. To do this we will look at the amount ribosomal and mitochonrial genes per sample and setup cutoff values for these.

Start with mitochondrial genes which are labelled starting with "MT"
```{r}
# generate function to calculate percent MT for each sample
do.MT <- function(obj) {
  obj[['percent.mt']] <- PercentageFeatureSet(obj, pattern = "^MT");
  obj
}
```

```{r}
#create list of objects
seur_obj <- list()
seur_obj[["RLN02"]] <- RLN02_seur
seur_obj[["RLN03"]] <- RLN03_seur
seur_obj[["RLN04"]] <- RLN04_seur
seur_obj[["RLN05"]] <- RLN05_seur
seur_obj[["CCS0025"]] <- CCS0025_seur
seur_obj[["GE0539"]] <- GE0539_seur
seur_obj[["DLC0087"]] <- DLC0087_seur
seur_obj[["DLC0341"]] <- DLC0341_seur
seur_obj[["GE0287"]] <- GE0287_seur
seur_obj[["DLC0169"]] <- DLC0169_seur
seur_obj[["CLC00277"]] <- CLC00277_seur
seur_obj[["GE0505"]] <- GE0505_seur
seur_obj[["CLC00275"]] <-CLC00275_seur
seur_obj[["CCS0099"]] <-CCS0099_seur
seur_obj[["CCS0277"]] <-CCS0277_seur
seur_obj[["CCS2010"]] <-CCS2010_seur

#run each sample through the mitochondrial function
seur_obj <- lapply(seur_obj, do.MT)
```

Now we can do the same thing for ribosomal genes which are denoted by "^RP[SL][[:digit:]]|^RP[[:digit:]]|^RPSA"
```{r}
# generate function to calculate percent ribosomes for each sample
do.RBP <- function(obj) {
  obj[['percent.rbp']] <- PercentageFeatureSet(obj, pattern = "^RP[SL][[:digit:]]|^RP[[:digit:]]|^RPSA");
  obj
}

#run each sample through the Ribosome function
seur_obj <- lapply(seur_obj, do.RBP)
```

Now we want to visualize these so we can setup our cutoff values for these
```{r}
#generate function to print volcano plot for each seurat object
do.VlnPlot <- function(obj){
  print(VlnPlot(obj, features = c("nFeature_RNA","nCount_RNA","percent.mt","percent.rbp"), ncol = 4))}

#run each seurat object through this new function
lapply(seur_obj, do.VlnPlot)
```

Based on these graphs we will filter out cells with nfeature count of <500 and filter out cells with a mitochondrial % of >25

```{r}
do.subset <- function(obj){
  obj <- subset(obj, subset = nFeature_RNA > 500 & percent.mt < 25)}

seur_obj <- lapply(seur_obj, do.subset)
```

Merge data together into a combined seurat object:
```{r}
#merge list
combined_seurat <- merge(x= seur_obj[[1]], y = seur_obj[-1])
```

Apply SCT transform to remove technical variation:
```{r}
combined_seurat <- SCTransform(combined_seurat, vars.to.regress = c("percent.mt","percent.rbp"))

#Remove extra datasets to free up space:
rm(seur_obj)
rm(RLN02_seur)
rm(RLN03_seur)
rm(RLN04_seur)
rm(RLN05_seur)
rm(GE0505_seur)
rm(CLC00277_seur)
rm(DLC0169_seur)
rm(GE0287_seur)
rm(DLC0341_seur)
rm(DLC0087_seur)
rm(GE0539_seur)
rm(CCS0025_seur)
rm(CLC00275_seur)
rm(CCS0099_seur)
rm(CCS0277_seur)
rm(CCS2010_seur)
```

## Integration of Datasets using Seurat integration method (v5)

Perform standard PCA and UMAP analysis on un-integrated cells
```{r}
set.seed(101)
combined_seurat <- RunPCA(combined_seurat, npcs = 30, verbose = F)
combined_seurat <- RunUMAP(combined_seurat, reduction = "pca", dims = 1:12, verbose = F)
```

Visualize UMAP before integration (Supplementary Figure S4A)
```{r int17,fig.align="center"}
DimPlot(combined_seurat,reduction = "umap", cols = c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#EE4C97", "#6A6767","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7")) + plot_annotation(title = "Before integration")

```

Integrate data:
```{r}
combined_seurat <- IntegrateLayers(
  object = combined_seurat,
  method = RPCAIntegration,
  new.reduction = "integrated.rpca",
  normalization.method = "SCT",
  verbose = F
)
```

```{r}
# perform UMAP analysis on integrated layer
set.seed(101)
combined_seurat <- RunUMAP(combined_seurat, reduction = "integrated.rpca", dims = 1:15, reduction.name = "umap.rpca")

```

Visualize UMAP after integration (Supplementary Figure S4B)
```{r int17,fig.align="center"}
DimPlot(combined_seurat, reduction = "umap.rpca", cols = c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#EE4C97", "#6A6767","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7")) + plot_annotation(title = "After integration")
```

```{r}
# rejoin layers
combined_seurat[["RNA"]] <- JoinLayers(combined_seurat[["RNA"]])
```

Save integrated and normalized seurat file
```{r}
saveRDS(combined_seurat, file = "~/trms_dlbcl/internal_data/combined_seurat.rds")
#combined_seurat <- readRDS("~/trms_dlbcl/internal_data/combined_seurat.rds")
```

# Find T cells

Use SingleR package to label immune cells for identification of T cells.

```{r}
DefaultAssay(combined_seurat) <- "RNA"
combined_seurat <- NormalizeData(combined_seurat)
combined_seurat <- FindVariableFeatures(combined_seurat, selection.method = "vst", nfeatures = 3000)
all.genes <- rownames(combined_seurat)
combined_seurat <- ScaleData(combined_seurat, features = all.genes)
```

Generate UMAP with labels:
```{r}

DimPlot(combined_seurat, cols = c("#BC3C29","#0072B5", "#E18727", "#20854E", "#7876B1", "#4b7284", "#ffc445", "#EE4C97", "#6A6767","#da6857","#7fb8da", "#EAAC6A", "#79B594", "#A8A7CD", "#6F99AD", "#FFDC91", "#f592c0", "#C5C6C7"))
```

**Labelling with Monaco Immune Data:**

```{r}
#get reference of Monaco Immune Data
monaco.ref <- celldex::MonacoImmuneData()

#convert seurat object to sce object
combined_sce <- as.SingleCellExperiment(DietSeurat(combined_seurat))

#SingleR functions to label cells
monaco.main <- SingleR(test = combined_sce,assay.type.test = 1,ref = monaco.ref,labels = monaco.ref$label.main)
```


Table of main labels:
```{r}
table(monaco.main$pruned.labels)
```

Add to labels to Seurat object:
```{r}
combined_seurat@meta.data$monaco.main <- monaco.main$pruned.labels
```

Generate UMAP with labels:
```{r}
set.seed(123)
combined_seurat <- SetIdent(combined_seurat, value = "monaco.main")

DimPlot(combined_seurat, label = T , repel = T, label.size = 3) + NoLegend()
```

**Labelling with Immune Cell Database**
```{r}
#get reference
dice.ref <- celldex::DatabaseImmuneCellExpressionData()

#SingleR functions to label cells
dice.main <- SingleR(test = combined_sce,assay.type.test = 1,ref = dice.ref,labels = dice.ref$label.main)
```

Table of main labels:
```{r}
table(dice.main$pruned.labels)
```

Add to labels to Seurat object:
```{r}
combined_seurat@meta.data$dice.main <- dice.main$pruned.labels
```

Generate UMAP with labels: (Supplementary Figure S4C)
```{r}
set.seed(123)
combined_seurat <- SetIdent(combined_seurat, value = "dice.main")
#Generate UMAP for supplemental Figure 5A
DimPlot(combined_seurat, label = T , repel = T, label.size = 3) + NoLegend()
```

Save labelled seurat
```{r}
saveRDS(combined_seurat, file = "~/trms_dlbcl/combined_seurat_label.rds")
```

## Subsetting T cells
Subset T cell cluster using DICE dataset
```{r}
set.seed(123)
#make sure ident is set to DICE dataset
combined_seurat <- SetIdent(combined_seurat, value = "dice.main")

#subset cluster of cells labelled as T cells and NK cells
all.Tcells <- subset(x=combined_seurat, idents = c("T cells, CD8+", "T cells, CD4+", "NK cells"))

```

PCA & Dimensionality Reduction of T/NK cells

```{r}
DefaultAssay(all.Tcells) <- "SCT"
all.Tcells <- RunPCA(all.Tcells, verbose = F)
all.Tcells <- RunUMAP(all.Tcells, dims = 1:20, verbose = F)
all.Tcells <- FindNeighbors(all.Tcells, dims = 1:20, k.param = 15, verbose = F)
all.Tcells <- FindClusters(all.Tcells, verbose = F)
table(all.Tcells[[]]$seurat_clusters)

#view plot
set.seed(123)
DimPlot(all.Tcells, label = T)
```

Remove NK cells by removing cells that are positive for NCAM1
```{r}
#first switch active assay to RNA
DefaultAssay(all.Tcells) <- "RNA"

# create new column in the metadata for Neg/Pos status of NCAM1
all.Tcells@meta.data$NK <- 'N'
all.Tcells@meta.data$NK[which(all.Tcells@assays$RNA$counts["NCAM1",]>0.1)] <- 'Y'

#set ident to NK
all.Tcells <- SetIdent(all.Tcells, value = "NK")

#subst NK cells that are positive for NCAM1
NK.cells <- subset(x=all.Tcells, idents = "Y")

#subset T cells that are negative for NCAM1
all.Tcells <- subset(x=all.Tcells, idents = c("N"))

```

## Determing Optimal clusters

Remove TCR genes for better clustering of cell phenotype
```{r}
DefaultAssay(all.Tcells) <- "SCT"

all.Tcells <- all.Tcells[!grepl("^TR[ABDG][VJC]", rownames(all.Tcells)), ]
```

Remove Ribosomal genes for better clustering
```{r}
all.Tcells <- all.Tcells[!grepl("^RP[SL][[:digit:]]|^RP[[:digit:]]|^RPSA", rownames(all.Tcells)), ]
```

Perform PCA and UMAP reduction on all T cells seurat object
```{r}
DefaultAssay(all.Tcells) <- "SCT"
all.Tcells <- RunPCA(all.Tcells, verbose = F)
all.Tcells <- RunUMAP(all.Tcells, dims = 1:16, verbose = F)
```

**Determining optimal number of clusters**

First try k-values of 10, 15, 20. 30
```{r}
#k=10
    all.Tcells <- FindNeighbors(all.Tcells, dims = 1:16, k.param = 10)
    all.Tcells<-FindClusters(all.Tcells, resolution=0.8)
    all.Tcells@meta.data$SCT_res0.8_k10 <- all.Tcells@meta.data$SCT_snn_res.0.8
    table(all.Tcells@meta.data$seurat_clusters)
    
#k=15
    all.Tcells <- FindNeighbors(all.Tcells, dims = 1:16, k.param = 15)
    all.Tcells<-FindClusters(all.Tcells, resolution=0.8)
    all.Tcells@meta.data$SCT_res0.8_k15 <- all.Tcells@meta.data$SCT_snn_res.0.8
    table(all.Tcells@meta.data$seurat_clusters)
#k=20
    all.Tcells <- FindNeighbors(all.Tcells, dims = 1:16, k.param = 20)
    all.Tcells<-FindClusters(all.Tcells, resolution=0.8)
    all.Tcells@meta.data$SCT_res0.8_k20 <- all.Tcells@meta.data$SCT_snn_res.0.8
    table(all.Tcells@meta.data$seurat_clusters)
    
    
#k=30
    all.Tcells <- FindNeighbors(all.Tcells, dims = 1:16, k.param = 30)
    all.Tcells<-FindClusters(all.Tcells, resolution=0.8)
    all.Tcells@meta.data$SCT_res0.8_k30 <- all.Tcells@meta.data$SCT_snn_res.0.8
    table(all.Tcells@meta.data$seurat_clusters)


```

Calculate Siloutte Index for each k-value

```{r}
require(cluster)
clusterings <- grepl("SCT_res0.8",colnames(all.Tcells@meta.data))
dist.matrix <- dist(x = Embeddings(object = all.Tcells[["pca"]])[, 1:16])
clusterings <- colnames(all.Tcells@meta.data)[clusterings]
for (c in clusterings) {
  require(cluster) # contains silhouette function.
  sil <- silhouette(as.numeric(as.character(all.Tcells@meta.data[,c])), dist=dist.matrix)
   print(paste(c))
  print(mean(sil[,3]))
  print(head(sil, n=20))
}
```

Decided to go ahead with k = 30 so now will try different resolutions to fine tune cluster number
```{r}

all.Tcells <- FindNeighbors(all.Tcells, dims = 1:16, k.param = 30)

for(res in c(0.2, 0.5, 0.6, 0.8, 1, 1.5)) {
    all.Tcells<-FindClusters(all.Tcells, resolution=res)
    print(paste(res))
    print(table(all.Tcells@meta.data$seurat_clusters))
}

```

Calculate Siloutte Index for each resolution:

```{r}
clusterings <- grepl("SCT_snn_res",colnames(all.Tcells@meta.data))
dist.matrix <- dist(x = Embeddings(object = all.Tcells[["pca"]])[, 1:16])
clusterings <- colnames(all.Tcells@meta.data)[clusterings]
for (c in clusterings) {
  require(cluster) # contains silhouette function.
  sil <- silhouette(as.numeric(as.character(all.Tcells@meta.data[,c])), dist=dist.matrix)
   print(paste(c))
  print(mean(sil[,3]))
  print(head(sil, n=20))
}
```

Calculate Rand Index for each resolution:
```{r}
require(igraph)

for (c in clusterings) {
  for (c2 in clusterings) {
    ARI <- compare(all.Tcells@meta.data[,c], all.Tcells@meta.data[,c2], method="adjusted.rand")
    print(paste(c, c2, ARI))
  }
}

```

Determine optimal clusters by using number of Marker genes per cluster:
```{r}
#prep for find markers
all.Tcells <- PrepSCTFindMarkers(object = all.Tcells, assay = "SCT", verbose = TRUE)
for (c in clusterings) {
  markers <- FindAllMarkers(all.Tcells, group.by=c, logfc.threshold = -Inf, only.pos=TRUE, max.cells.per.ident=100, verbose=FALSE)
  markers <- markers[ markers[,"p_val_adj"] < 0.05, ]
  print(paste(c))
  print(table(markers$cluster))
}

```

Resolution of 0.8 looks best in terms of marker genes so will go ahead with K = 30 and resolution = 0.8

```{r}
all.Tcells <- FindNeighbors(all.Tcells, dims = 1:16, k.param = 30)
all.Tcells<-FindClusters(all.Tcells, resolution=0.8)

# vislualize UMAP of clusters
DimPlot(all.Tcells, label = T)

#save seurat object of clustered T cells
saveRDS(all.Tcells, file = "~/trms_dlbcl/all_T_clustered.rds")
```

## Labelling of other cell types on all cells UMAP

*Label Monocytes/DCs:*
Subset cells labelled as Monocytes by DICE database
```{r}
combined_seurat <- SetIdent(combined_seurat, value = "dice.main")
M.DC.cells <- subset(x=combined_seurat, idents = "Monocytes")

M.DC.cells@meta.data$cellID <- "Monocytes"
```

Decipher between monocytes and DCs by using LYZ and LILRA4 genes to determine which cells are DCs
```{r}
# create new column in the metadata for Neg/Pos status of gene for DC calls (NCAM1)
M.DC.cells@meta.data$DC <- 'N'
M.DC.cells@meta.data$DC[which(M.DC.cells@assays$RNA$counts["LYZ",]>0)] <- 'Y'
M.DC.cells@meta.data$DC[which(M.DC.cells@assays$RNA$counts["LILRA4",]>0)] <- 'Y'

#make sure ident is set to DC ident
M.DC.cells <- SetIdent(M.DC.cells, value = "DC")

#subset DCs
DCs <- subset(x=M.DC.cells, idents = c("Y"))

#create cellID label for DCs
DCs@meta.data$cellID <- "DC"

#create cellID label for NK cells
NK.cells@meta.data$cellID <- "NK"
```


*Determining B cell status*
Add metadata to seurat object so have classification if cell is from rLN or DLBCL patient. 
```{r}
#load new metadata
md <- read.csv("~/trms_dlbcl/internal_data/metadata.csv", stringsAsFactors = FALSE)

#create new df joining md to seurat samples based on original sample ID
add_data <- left_join(combined_seurat[["orig.ident"]], md)

#covert df to have rownames of cell ids
row.names(add_data) <- row.names(combined_seurat[[]])

#finally add metadata to seurat object
combined_seurat <- AddMetaData(combined_seurat, metadata = add_data)

```

Subset B cells:

```{r}
combined_seurat <- SetIdent(combined_seurat, value = "dice.main")

B.cells <- subset(x=combined_seurat, idents = "B cells")
```

Cluster B cells

```{r}
DefaultAssay(B.cells) <- "SCT"
B.cells <- RunPCA(B.cells, verbose = F)
B.cells <- RunUMAP(B.cells, dims = 1:30, verbose = F)
B.cells <- FindNeighbors(B.cells, dims = 1:30, k.param = 20, verbose = F)
B.cells <- FindClusters(B.cells, verbose = F, resolution=0.8)
table(B.cells[[]]$seurat_clusters)

#view output of clusters
set.seed(123)
DimPlot(B.cells, label = T)
```


Look for IGKC expression per B cell cluster
```{r}
B.cells <- SetIdent(B.cells, value= "Control_or_lymphoma")
DimPlot(B.cells)
FeaturePlot(B.cells, features = "IGKC")
B.cells <- SetIdent(B.cells, value= "seurat_clusters")
VlnPlot(B.cells, features = "IGKC")
```

Ratio of Control or lymphoma per B cell cluster
```{r}

dittoBarPlot(B.cells, "Control_or_lymphoma",group.by = "seurat_clusters")

```

Now add a new column of wether each is lymphoma or not as derermined by which clusters are made up of only (or mostly) DLBCL patient cells and have low IGKC expression
```{r}
#get number of clusters
nclusters <- length(unique(B.cells@meta.data$seurat_clusters))
# Create a vector where the names of the vector are the cluster number
# and the values of the vector are the annotated names.
cluster_annotation2 <- rep("unannotated", nclusters)
names(cluster_annotation2) <- levels(B.cells@meta.data$seurat_clusters)
cluster_annotation2["0"] = "Malignant"
cluster_annotation2["1"] = "Malignant"
cluster_annotation2["2"] = "Normal"
cluster_annotation2["3"] = "Malignant"
cluster_annotation2["4"] = "Malignant"
cluster_annotation2["5"] = "Malignant"
cluster_annotation2["6"] = "Malignant"
cluster_annotation2["7"] = "Malignant"
cluster_annotation2["8"] = "Malignant"
cluster_annotation2["9"] = "Malignant"
cluster_annotation2["10"] = "Malignant"
cluster_annotation2["11"] = "Normal"
cluster_annotation2["12"] = "Malignant"
cluster_annotation2["13"] = "Normal"
cluster_annotation2["14"] = "Malignant"
cluster_annotation2["15"] = "Malignant"
cluster_annotation2["16"] = "Normal"
cluster_annotation2["17"] = "Malignant"
cluster_annotation2["18"] = "Normal"
cluster_annotation2["19"] = "Malignant"
cluster_annotation2["20"] = "Normal"
cluster_annotation2["21"] = "Malignant"
cluster_annotation2["22"] = "Normal"
cluster_annotation2["23"] = "Normal"
cluster_annotation2["24"] = "Malignant"
cluster_annotation2["25"] = "Malignant"
cluster_annotation2["26"] = "Malignant"
cluster_annotation2["27"] = "Malignant"
cluster_annotation2["28"] = "Malignant"
cluster_annotation2["29"] = "Normal"
cluster_annotation2["30"] = "Normal"
cluster_annotation2["31"] = "Malignant"
cluster_annotation2["32"] = "Malignant"
cluster_annotation2["33"] = "Malignant"
cluster_annotation2["34"] = "Normal"

B.cells <- SetIdent(B.cells, value=B.cells@meta.data$seurat_clusters)
B.cells <- RenameIdents(B.cells, cluster_annotation2)
B.cells@meta.data$cellID <- Idents(B.cells)
head(B.cells@meta.data)
```

Check to make sure B cells from rLN samples are not being classified as malignant B cells
```{r}
dittoBarPlot(B.cells, "Control_or_lymphoma",group.by = "cellID")

```

Very few rLN B cells were classified as malignant so we are ok with this classification.

*Label T cell phenotypes*
Now add a new column of general T cell phenotypes determined in internal_TcellAnalysis.rmd
```{r}
#get number of clusters
nclusters <- length(unique(all.Tcells@meta.data$seurat_clusters))
# Create a vector where the names of the vector are the cluster number
# and the values of the vector are the annotated names.
cluster_annotation2 <- rep("unannotated", nclusters)
names(cluster_annotation2) <- levels(all.Tcells@meta.data$seurat_clusters)
cluster_annotation2["0"] = "Naive"
cluster_annotation2["1"] = "Effector"
cluster_annotation2["2"] = "Tfh"
cluster_annotation2["3"] = "Memory"
cluster_annotation2["4"] = "Memory"
cluster_annotation2["5"] = "Naive"
cluster_annotation2["6"] = "Effector"
cluster_annotation2["7"] = "Effector"
cluster_annotation2["8"] = "Treg"
cluster_annotation2["9"] = "Effector"
cluster_annotation2["10"] = "Treg"
cluster_annotation2["11"] = "Effector"
cluster_annotation2["12"] = "Tfh"
cluster_annotation2["13"] = "Memory"
cluster_annotation2["14"] = "Memory"
cluster_annotation2["15"] = "Proliferating"
cluster_annotation2["16"] = "Naive"
cluster_annotation2["17"] = "IFN-signaling"

all.Tcells <- SetIdent(all.Tcells, value=all.Tcells@meta.data$seurat_clusters)
all.Tcells <- RenameIdents(all.Tcells, cluster_annotation2)
all.Tcells@meta.data$cellID <- Idents(all.Tcells)
head(all.Tcells@meta.data)
```

Now add new cellID labels to combined_seurat object
```{r}
#label all cells as other to start
combined_seurat$cellID <- "other"
#add B cells to combined_seurat object
combined_seurat@meta.data[Cells(B.cells), "cellID"] <-as.character(B.cells$cellID)
#add Monocytes/DCs to combined_seurat object
combined_seurat@meta.data[Cells(M.DC.cells), "cellID"] <- M.DC.cells$cellID
combined_seurat@meta.data[Cells(DCs), "cellID"] <- DCs$cellID
#add NK to combined_seurat object
combined_seurat@meta.data[Cells(NK.cells), "cellID"] <- NK.cells$cellID
#add T cells to combined_seurat object
combined_seurat@meta.data[Cells(all.Tcells), "cellID"] <- as.character(all.Tcells$cellID)

head(combined_seurat@meta.data)
```

Generate UMAP with label of if cells came from DLBCL or rLN samples (Supplementary Figure S4D)
```{r}
combined_seurat <- SetIdent(combined_seurat, value = "Control_or_lymphoma")
DimPlot(combined_seurat, cols = c("#6F99AD", "#BC3C29"))
```

Generate UMAP with new cell labels (Supplementary Fig S4E)
```{r}
combined_seurat <- SetIdent(combined_seurat, value = "cellID")
DimPlot(combined_seurat, label = T)
```

## Look at expression of factors known to promote TRM cell generation on all cells from TME
Will look at expression of: 
* E-cadherin (CDH1) - ligand for CD103
* TGF-beta (TGFB1) - induces expression of CD103
* Retinol dehydrogenase 10 (RDH10) and Aldehyde dehydrogenase family members (ALDH1A1, ALDH1A2, ALDH1A3) which is important in synthesis of retonic acid which has been found to play a role in TRM formation

Look at expression of these genes in all cells between RLN vs DLBCL patients (Supplementary Fig. S5C)
```{r}
combined_seurat <- SetIdent(combined_seurat, value = "Control_or_lymphoma")
test_sign = list(c("LBCL", "RLN"))

VlnPlot(object = combined_seurat, features = c('CDH1','TGFB1', 'RDH10', 'ALDH1A1', 'ALDH1A2', 'ALDH1A3'),  cols = c("#6F99AD", "#BC3C29"), pt.size = 0.02, add.noise = F)+ stat_compare_means(comparisons = test_sign, label.y = 3)

```

Look at expression of these genes in different cell types in RLN vs DLBCL patients (Supplementary Fig. S5D)
```{r}
combined_seurat <- SetIdent(combined_seurat, value = "cellID")

#order cells
combined_seurat$cellID <- factor(x = combined_seurat$cellID, levels = c('Malignant', 'Normal', "Monocytes", "DC", "NK", "Treg", "Tfh", "Naive", "Effector", "Memory", "Proliferating", "IFN-signaling", "other"))

#plot
pdf("~/trms_dlbcl/internal_data/tme_vln.pdf", height = 8, width = 12)
VlnPlot(object = combined_seurat, features = c('CDH1','TGFB1', 'RDH10', 'ALDH1A1', 'ALDH1A2', 'ALDH1A3'), split.by = "Control_or_lymphoma", cols = c("#BC3C29","#6F99AD"), group.by = "cellID")
dev.off()
```
